Specialisation in food webs scales with species richness but not with latitude

### 2015-07-01 manuscript/latitudepaper.tex
Version uploaded to Proc R Soc

  If the latitude-niche breadth hypothesis is correct, there should also be direct relationships between latitude and the
  degree of specialisation (i.e., the breadth of the Eltonian niche ~\citep{Elton1927}) of species within food webs.
  Attempts to unravel these effects, however,
  are complicated by known relationships between species richness and many other network properties~\citep{Riede2010}. 
  For example,
  narrower niches imply fewer links per species (i.e., greater specialisation) in the tropics(~\citealp{Marra1997,Dyer2007};
  but see~\citealp{Schleuning2012}). However,
  average numbers of links per species tend to increase in larger food webs~\citep{Dunne2006,Riede2010}. This means
  that a latitudinal effect on specialisation may be obscured by a latitudinal gradient in species richness. If this is
  the case, it may still be possible to uncover effects of latitude on specialisation by examining the shape of the scaling 
  relationship between specialisation and species richness over changing latitude. Here, we use three measures of specialisation:
  mean links per species, mean generality (number of prey), and mean vulnerability (number of predators). By testing whether
  latitude affects the scaling of each property with species richness, we test effects of latitude on specialisation implied by
  the ``latitude-niche breadth hypothesis''.

@@

    We compiled a list of 163 empirical food webs from
    multiple sources (see \emph{Appendix S1} for web origins and selection
    criteria). We recorded study site latitude from the original source where
    possible or, where study sites were described but exact coordinates were not
    given, obtained estimated coordinates using Google Earth~\citep{GoogleEarth}.
    We then divided the species in each web into basal resources (those species with
    consumers but no prey), top predators (those species with prey but no predators),
    and intermediate consumers (species with both predators and prey, including cannibalistic
    species).

@@

    \noindent where $S_{i}$ is the species richness of web $i$, $L_{i}$ its absolute
    latitude (degrees north or south  regardless of direction), $E_{i}$ is a categorical
    variable indicating the ecosystem type of network $i$ (comprising terms for stream, 
    marine, lake, and terrestrial networks with estuarine
    networks providing the intercept) and $\epsilon_{i}$ a residual error term.
    The logarithmic formulation of this model is

@@

  Unlike species richness, only generality showed an overall increase with
  increasing proportions of basal resources in a web ($\beta_0$=0.019,
  $p$=0.859; $\beta_0$=0.383, $p$=0.001; and $\beta_0$=0.019, $p$=0.855 for
  link density, generality, and vulnerability, respectively;
  Fig.~\ref{props_v_lat}). In each case, however, there were substantial
  effects of latitude on the scaling relationships. Thus, the scaling
  exponents for link density and vulnerability were negative for estuarine,
  lake, and stream communities and became more negative towards the poles
  ($\beta_{Latitude}$=-0.008, $p$\textless0.001 and $\beta_{Latitude}$=-0.008,
  $p$\textless0.001, respectively; Fig.~\ref{BIT}). For the scaling of link
  density and vulnerability in marine and terrestrial communities, the overall
  effect of latitude on scaling was positive ($\beta_{Latitude:Marine}$=0.011,
  $p$=0.020; $\beta_{Latitude:Terrestrial}$=0.016, $p$\textless0.001 and
  $\beta_{Latitude:Marine}$=0.011, $p$=0.020;
  $\beta_{Latitude:Terrestrial}$=0.016, $p$\textless0.001, respectively) such
  that there was little overall scaling relationship. In contrast, the scaling
  exponent of generality with the proportion of basal resources was positive
  near the equator and negative near the poles in estuarine, lake, and stream
  communities ($\beta_{Latitude}$=-0.010, $p$\textless0.001) while there was
  very little change in the scaling exponents for marine and terrestrial
  communities ($\beta_{Latitude:Marine}$=0.013, $p$=0.010;
  $\beta_{Latitude:Terrestrial}$=0.019, $p$\textless0.001).

@@

  In contrast to the generally positive scaling relationships above, link
  density, generality, and vulnerability decreased as the proportion of top
  predators increased ($\beta_0$=-0.532, $p$\textless0.001; $\beta_0$=-0.454,
  $p$\textless0.001; and $\beta_0$=-0.532, $p$\textless0.001 respectively;
  Fig~\ref{props_v_lat}). For estuarine, lake, and marine food webs the
  strength of this scaling did not vary significantly with latitude
  ($\beta_{Latitude}$=-0.001, $p$=0.481 for link density;
  $\beta_{Latitude}$=-0.001, $p$=0.472 for vulnerability) except in stream and terrestrial food
  webs, where scaling was more strongly negative in the tropics and near zero
  near the poles ($\beta_{Latitude:Stream}$=0.012, $p$=0.001 and $\beta_{Latitude:Terrestrial}$=0.009, $p$=0.015 for link density;
  $\beta_{Latitude:Stream}$=0.012, $p$=0.001 and
  $\beta_{Latitude:Terrestrial}$=0.009, $p$=0.015
  for vulnerability;
  Fig~\ref{BIT}). The best-fit model for generality did not include any terms
  for latitude in any ecosystem type.

@@

  The tendency of food-web structure to exhibit scaling relationships with
  species richness has been well-established~\citep{Dunne2004,Riede2010}. As
  species richness in particular is also known to vary systematically over
  latitude~\citep{Kaufman1995,Schemske2009,Macpherson2002}, intuitively one
  might suspect that any relationship between food-web properties such as
  generality might be due to the latitudinal gradient in species richness. In
  this dataset, however, we did not find overall latitudinal gradients in
  species richness, links per species,  generality, vulnerability, or the
  proportions of food webs accounted for by  basal resources, intermediate
  consumers, and top predators (see \emph{Appendix S3} for details).

  The lack of a latitudinal gradient in species richness in this dataset
  contrasts strongly with other
  studies~\citep{Kaufman1995,Schemske2009,Macpherson2002}. As numbers of
  species and links included in a food web vary strongly with sampling effort
  as well as with the underlying diversity of the study area, it is possible
  that the lack of latitudinal trends here is a result of researchers tending
  to expend similar amounts of sampling effort across studies. This could
  result in food webs describing species-rich tropical communities omitting
  more species and links than species-poor arctic communities. In addition, it
  is worth noting that gradients in species richness are generally measured
  for a single taxonomic group at a
  time~\citep[e.g.]{Kaufman1995,Schemske2009,Macpherson2002}. It is possible
  that these taxa are not well-represented in our food webs and that the
  dominant taxa in them do not have an underlying latitudinal gradient in
  richness.In either case, the lack of a strong association between species
  richness and latitude in any ecosystem type means that any effect of
  latitude on other scalingrelationships is not being driven by an
  underlyinglatitudinal gradient in species richness. The lack of confounding
  effects of latitude allows us tomore clearly assess effects of latitude on
  scaling with species richness and proportions of species in differenttrophic
  levels.

@@

  The scaling of link density, generality and vulnerability with proportions
  of species at different trophic levels, however, did not differ between lakes
  and most other food webs. Indeed, there was comparatively little variation in
  scaling with the trophic-level breakdown of a web across ecosystem types.
  Terrestrial food webs tended to exhibit less change in scaling over latitude
  than other ecosystem types, as did marine webs when considering scaling with
  the proportion of basal resources. More excitingly, the negative relationships
  between the proportion of top predators in a web and link density and
  vulnerability were weaker (i.e., the scaling exponent was more strongly
  negative) in low-latitude stream ecosystems. This may imply that top predators
  in high-latitude streams tend to be more generalist than their low-latitude
  equivalents~\citep{Winemiller2008}, perhaps due to prey-switching during seasonal
  food shortages~\citep{Magalhaes1993}.
  More generally, the lack of correspondence between the scaling of food-web
  properties with species richness and with proportions of species at each
  trophic level suggests that the size and trophic breakdowns of a community can
  each provide different information~\citep{Downing2002}.

@@

\section*{Figures}
\begin{figure}[h]
\centerline{\includegraphics*[width=.85\textwidth]{Figures/by_TL/scaling_with_S/proportions/fitlines_nonts.eps}}
\caption{Scaling relationships for re-scaled link density, generality, and vulnerability 
relative to species richness and the proportions of basal resources (\% Basal), intermediate 
consumers (\% Intermediate), or top predators (\% Top) in a food web. Link density, generality,
and vulnerability were each re-scaled to remove the effects of latitude and ecosystem
type. As the these relationships take the form of power laws, we did this by dividing the food-web
property (e.g. link density) by the predictor (e.g. \% Basal) raised to an exponent including the 
effects of latitude and, where applicable, ecosystem type and the interaction between ecosystem
type and latitude. Note that in all cases estuarine food webs were used as the baseline 
ecosystem type, but that at most two ecosystem types had interactions between ecosystem type and
latitude retained in the best-fit model (see \emph{results} for specifics). For each relationship, 
we show the rescaled values (white circles) as well as the overall scaling relationship using estuarine
ecosystems as a baseline (black line, N=163 food webs). For a figure with the uncorrected values,
see Fig.~\emph{S1}, \emph{Appendix S4}.}
\label{props_v_lat}
\end{figure}

\begin{figure}[h]
\centerline{\includegraphics*[height=.65\textheight]{Figures/by_TL/marginal/justS_marginal_latitude.eps}}
\caption{Changes to the scaling of link density with species richness across ecosystem
types and over latitude. We show the estimated scaling exponent for species richness (black
line) with its 95\% confidence interval (in grey), based on N=163 empirical food webs.
Latitude is given in degrees from the equator regardless of direction. The
behaviour of exponents for the scaling of generality and vulnerability with
species richness was very  similar to those for link density, except for the
scaling of generality in streams where  the size of the exponent increased
towards the poles. See Fig.~\emph{S2, Appendix 5} for  all scaling
relationships.} \label{S} \end{figure}

\begin{figure}[!H]
\centerline{\includegraphics*[height=.65\textheight]{Figures/by_TL/marginal/BIT_marginal_latitude.eps}}
\caption{Changes to the scaling of link density with the proportions of basal resources, intermediate
consumers, or top predators in a food web across ecosystem types and over latitude. For each proportion
we show the estimated scaling exponent (black line) with its 95\% confidence interval (in grey), based
on N=163 empirical food webs. Latitude is given in degrees from the equator regardless of direction. 
The behaviour of exponents for the scaling relationships of generality and vulnerability with each 
proportion was very similar to those of the scaling relationships with link density, except for the 
scaling of generality with the proportion of top predators where there was no effect of latitude on the 
size of the exponent in any ecosystem type. See Figs.~\emph{S3-S5, Appendix 5} for all scaling relationships.}
\label{BIT}
\end{figure}

>>>>

  If the latitude-niche breadth hypothesis is correct, there should also be direct relationships between latitude and the
  degree of specialisation (i.e., the breadth of the Eltonian niche; ~\citealp{Elton1927}) of species within food webs.
  Attempts to unravel these effects, however,
  are complicated by known relationships between species richness and many other network properties~\citep{Riede2010}. 
  For example,
  narrower niches imply fewer links per species (i.e., greater specialisation) in the tropics (\citealp{Marra1997,Dyer2007};
  but see~\citealp{Schleuning2012}). However,
  average numbers of links per species tend to increase in larger food webs~\citep{Dunne2006,Riede2010}. This means
  that a latitudinal effect on specialisation may be obscured by a latitudinal gradient in species richness. If this is
  the case, it may still be possible to uncover effects of latitude on specialisation by examining the shape of the scaling 
  relationship between specialisation and species richness over changing latitude. Here, we use three measures of specialisation:
  mean links per species, mean generality (number of prey), and mean vulnerability (number of predators). By testing whether
  latitude affects the scaling of each property with species richness, we test effects of latitude on specialisation implied by
  the ``latitude-niche breadth hypothesis''.

@@

    We compiled a list of 166 empirical food webs from
    multiple sources (see \emph{Appendix S1} for web origins and selection
    criteria). We recorded study site latitude from the original source where
    possible or, where study sites were described but exact coordinates were not
    given, obtained estimated coordinates using Google Earth~\citep{GoogleEarth}.
    We then divided the species in each web into basal resources (those species with
    consumers but no prey), top predators (those species with prey but no predators),
    and intermediate consumers (species with both predators and prey, including cannibalistic
    species).

@@

    \noindent where $S_{i}$ is the species richness of web $i$, $L_{i}$ its absolute
    latitude (degrees north or south  regardless of direction), $E_{i}$ is a categorical
    variable indicating the ecosystem type of network $i$ (comprising terms for stream, 
    marine, lake, and terrestrial networks with estuarine
    networks providing the intercept) and $\epsilon_{i}$ is a residual error term.
    The logarithmic formulation of this model is

@@

  Unlike species richness, only generality showed an overall increase with
  increasing proportions of basal resources in a web ($\beta_0$=0.019,
  $p$=0.859; $\beta_0$=0.383, $p$=0.001; and $\beta_0$=0.019, $p$=0.855 for
  link density, generality, and vulnerability, respectively;
  Fig.~\ref{props_v_lat}). In each case, however, there were substantial
  effects of latitude on the scaling relationships. Thus, the scaling
  exponents for link density and vulnerability were negative for estuarine,
  lake, and stream communities and became more negative towards the poles
  ($\beta_{Latitude}$=-0.008, $p$\textless0.001 and $\beta_{Latitude}$=-0.008,
  $p$\textless0.001, respectively; Fig.~\ref{BIT}). For the scaling of link
  density and vulnerability in marine and terrestrial communities, the overall
  effect of latitude on scaling was positive ($\beta_{Latitude:Marine}$=0.011,
  $p$=0.020; $\beta_{Latitude:Terrestrial}$=0.016, $p$\textless0.001 and
  $\beta_{Latitude:Marine}$=0.011, $p$=0.020;
  $\beta_{Latitude:Terrestrial}$=0.016, $p$\textless0.001, respectively) such
  that there was weak scaling at most latitudes. In contrast, the scaling
  exponent of generality with the proportion of basal resources was positive
  near the equator and negative near the poles in estuarine, lake, and stream
  communities ($\beta_{Latitude}$=-0.010, $p$\textless0.001) while there was
  very little change in the scaling exponents for marine and terrestrial
  communities ($\beta_{Latitude:Marine}$=0.013, $p$=0.010;
  $\beta_{Latitude:Terrestrial}$=0.019, $p$\textless0.001).

@@

  In contrast to the generally positive scaling relationships above, link
  density, generality, and vulnerability decreased as the proportion of top
  predators increased ($\beta_0$=-0.532, $p$\textless0.001; $\beta_0$=-0.454,
  $p$\textless0.001; and $\beta_0$=-0.532, $p$\textless0.001 respectively;
  Fig~\ref{props_v_lat}). For estuarine, lake, and marine food webs the
  strength of this scaling did not vary significantly with latitude
  ($\beta_{Latitude}$=-0.001, $p$=0.481 for link density;
  $\beta_{Latitude}$=-0.001, $p$=0.472 for vulnerability).
  In stream and terrestrial food webs, however, scaling was more strongly negative in the 
  tropics and near zero near the poles ($\beta_{Latitude:Stream}$=0.012, $p$=0.001 
  and $\beta_{Latitude:Terrestrial}$=0.009, $p$=0.015 for link density;
  $\beta_{Latitude:Stream}$=0.012, $p$=0.001 and $\beta_{Latitude:Terrestrial}$=0.009, $p$=0.015
  for vulnerability; Fig.~\ref{BIT}. The best-fit model for generality did not include any terms
  for latitude in any ecosystem type.

@@

  The tendency of food-web structure to exhibit scaling relationships with
  species richness has been well-established~\citep{Dunne2004,Riede2010}. As
  species richness in particular is also known to vary systematically over
  latitude~\citep{Kaufman1995,Macpherson2002,Hillebrand2004,Schemske2009},
  intuitively one might suspect that any relationship between food-web
  properties such as generality might be due to the latitudinal gradient in
  species richness. In this dataset, however, we did not find overall
  latitudinal gradients in species richness, links per species,generality,
  vulnerability, or the proportions of food webs accounted for by basal
  resources, intermediate consumers, and top predators, except in lake and
  stream food webs where the proportions of top predators tended to decrease
  towards the poles  (see \emph{Appendix S3} for details).

  The lack of a latitudinal gradient in species richness in this dataset
  contrasts strongly with other
  studies~\citep{Kaufman1995,Macpherson2002,Hillebrand2004,Schemske2009}. As
  numbers of species and links included in a food web vary strongly with
  sampling effort as well as with the underlying diversity of the study area,
  it is possible that the lack of latitudinal trends here is a result of
  researchers tending to expend similar amounts of sampling effort across
  studies. This could result in food webs describing species-rich tropical
  communities omitting more species and links than species-poor arctic
  communities. In addition, it is worth noting that gradients in species
  richness are generally measured for a single taxonomic group at a
  time~\citep{Kaufman1995,Macpherson2002,Hillebrand2004,Schemske2009}. It is possible
  that these taxa are not well-represented in our food webs and that the
  dominant taxa in them do not have an underlying latitudinal gradient in
  richness.In either case, the lack of a strong association between species
  richness and latitude in any ecosystem type means that any effect of
  latitude on other scaling relationships is not being driven by an underlying
  latitudinal gradient in species richness. The lack of confounding effects of
  latitude allows us to more clearly assess effects of latitude on scaling
  with species richness and proportions of species in different trophic
  levels.

@@

  The scaling of link density, generality and vulnerability with proportions
  of species at different trophic levels, however, did not differ between
  lakes and most other food webs. Indeed, there was comparatively little
  variation in scaling with the trophic-level breakdown of a web across
  ecosystem types. Nevertheless, the negative relationships between the
  proportion of top predators in a web and link density and vulnerability were
  weaker (i.e., the scaling exponent was more strongly negative) in low-latitude 
  stream ecosystems. This may imply that top predators in high-latitude 
  streams tend to be more generalist than their low-latitude
  equivalents~\citep{Winemiller2008}, perhaps due to prey-switching during
  seasonal food shortages~\citep{Magalhaes1993}. More generally, the lack of
  correspondence between the scaling of food-web properties with species
  richness and with proportions of species at each trophic level suggests that
  the size and trophic breakdowns of a community can each provide different
  information~\citep{Downing2002}.

@@

\section*{Figures}
\begin{figure}[h]
% \centerline{\includegraphics*[width=.85\textwidth]{Figures/by_TL/scaling_with_S/proportions/fitlines_nonts.eps}}
\caption{Scaling relationships for re-scaled link density, generality, and vulnerability 
relative to species richness and the proportions of basal resources (\% Basal), intermediate 
consumers (\% Intermediate), or top predators (\% Top) in a food web. Link density, generality,
and vulnerability were each re-scaled to remove the effects of latitude and ecosystem
type. As these relationships take the form of power laws, we did this by dividing the food-web
property (e.g. link density) by the predictor (e.g. \% Basal) raised to an exponent including the 
effects of latitude and, where applicable, ecosystem type and the interaction between ecosystem
type and latitude. Note that in all cases estuarine food webs were used as the baseline 
ecosystem type, but that at most two ecosystem types had interactions between ecosystem type and
latitude retained in the best-fit model (see \emph{results} for specifics). For each relationship, 
we show the re-scaled values (white circles) as well as the overall scaling relationship using estuarine
ecosystems as a baseline (black line, N=166 food webs). For a figure with the uncorrected values,
see Fig.~\emph{S1}, \emph{Appendix S4}.}
\label{props_v_lat}
\end{figure}

\begin{figure}[h]
% \centerline{\includegraphics*[height=.65\textheight]{Figures/by_TL/marginal/justS_marginal_latitude.eps}}
\caption{Changes to the scaling of link density with species richness across ecosystem
types and over latitude. We show the estimated scaling exponent for species richness (black
line) with its 95\% confidence interval (in grey), based on N=166 empirical food webs.
Latitude is given in degrees from the equator regardless of direction. The
behaviour of exponents for the scaling of generality and vulnerability with
species richness was very  similar to those for link density, except for the
scaling of generality in streams where  the size of the exponent increased
towards the poles. See Fig.~\emph{S2, Appendix 5} for  all scaling
relationships.} \label{S} \end{figure}

\begin{figure}[!H]
% \centerline{\includegraphics*[height=.65\textheight]{Figures/by_TL/marginal/BIT_marginal_latitude.eps}}
\caption{Changes to the scaling of link density with the proportions of basal resources, intermediate
consumers, or top predators in a food web across ecosystem types and over latitude. For each proportion
we show the estimated scaling exponent (black line) with its 95\% confidence interval (in grey), based
on N=166 empirical food webs. Latitude is given in degrees from the equator regardless of direction. 
The behaviour of exponents for the scaling relationships of generality and vulnerability with each 
proportion was very similar to those of the scaling relationships with link density, except for the 
scaling of generality with the proportion of top predators where there was no effect of latitude on the 
size of the exponent in any ecosystem type. See Figs.~\emph{S3-S5, Appendix 5} for all scaling relationships.}
\label{BIT}
\end{figure}
